### 2016-02-24 ch02-qlogms/excitation-design.tex
Improve CH02

\chapter{Design of excitation signals}
\label{sec:excitation}
\def\thisDir{ch02-qlogms}
\myEpigraph{A common mistake that people make when trying to design something completely foolproof is to underestimate the ingenuity of complete fools.}{Douglas Adams}{Mostly Harmless}
\researchBasedOn{Larsson2012SYSID, Geerardyn2012IMTC, Geerardyn2013TIM, }
\TODO{rename labels}

@@

Such problem settings occur in many diverse fields of experimentation, such as the measurement of biochemical impedances\citep{Bragos2001,Sanchez2011}, electrochemical impedance spectroscopy~\citep{Niedostatkiewicz2009,VanGheem2004}, electronics\TODO{citation}, mechatronics\TODO{citation} and vibration testing\TODO{citation (ISMA?)}.

@@

  \paragraph{Contents}
  In \secref{sec:excitation:inputDesign}, we first discuss (optimal) input design on an example second-order system.

  \TODO{
  \begin{itemize}
    \item input design
    \item system parametrization
    \item $1/f$ power
    \item frequency grid specification
    \item optimization of frequency grid
    \item simulations
    \item measurements
  \end{itemize}
  }

  \TODO{somewhere: mention D-optimal is indep of parametrization}
  \TODO{somewhere: mention optimality criterion is of lesser importance for a ``good'' signal (optimal in X = good in Y in most cases)}

  In the first section of this chapter, we start from a generic transfer function model and discuss a decomposition thereof in second order subsystems and their properties.
  We provide a comparison between several possible implementations of the frequency grids.
  We discuss their applicability to multisine signals.
  We also introduce a compensation method for the problems of the quasi-logarithmic grid and propose a novel $1/f$ `pink' quasi-logarithmic multisine.
  The second section describes the selection of a suitable frequency ratio $\alpha$ for a logarithmically spaced grid such that constant quality (signal to noise ratio (SNR)) is obtained.

  In the third section, the presented techniques are illustrated on a simulation example.
  Finally, we verify our results experimentally on the measurements of a band-pass filter.

\section{Input design}
\label{sec:excitation:inputDesign}
For the theoretical analysis, we consider \gls{SISO} \gls{OE} systems, i.e.
\begin{align}
y(t) &=G(s,\theta)u(t)  + e(t),
\end{align}
where $u(t)$ is a known input sequence, $y(t)$ the output sequence and $e(t)$ is zero mean Gaussian white noise.
$G$ is a rational, time invariant transfer function in continuous time. 
For notational simplicity, $s$ is used both as the complex Laplace variable and the differentiation operator; it should be clear from the context which is used. 
The transfer function is parameterized by the vector $\theta$.
These parameters are unknown and therefore need to be estimated.

@@

In open loop experiments, the scaled average information matrix is given by
\begin{align}
\IF & = \lim_{N\rightarrow\infty}
          \frac{1}{N}
            \sum_{t=1}^{N}
              {\left(\frac{\partial\hat{y}(t)}{\partial\theta}\right)
               \left(\frac{\partial\hat{y}(t)}{\partial\theta}\right)^{\T}},\\
\frac{\partial\hat{y}(t)}{\partial\theta}
    & =\frac{\partial G(s,\theta)}{\partial \theta}u(t).
\end{align}
By Parseval's theorem, $\IF$ can be expressed as
\begin{align}
\IF &= \frac{1}{\pi}
         \int\limits_0^\infty \mathrm{Re}
           \left\{      \frac{\partial G(j\omega)}{\partial \theta}
                  \left[\frac{\partial G(j\omega)}{\partial \theta}\right]^{\HT}
           \right\}
           \powerspec{u}(j\omega)
           \dd{\omega} \text{.}
\label{eq:excitation:IF}
\end{align}
The information matrix is a function of the input power spectrum $\powerspec{u}$, which is the only entity that can be used to improve the quality of the estimates. A common optimality criterion is D-optimality where $\det{\IF}$ is maximized \citep{Goodwin1977}.

@@

%Such a signal is discussed in the next section.

% \TODO{begin}
% \newcommand{\robust}{\mathrm{robust}\}
% \subsection{Using prior system information}
% Consider a system $G(s,\theta)$ where some initial guess of the parameter vector,
% say $\hat{\theta}_0 \in \mathbb{R}^n$ exists. A D-optimal design can then be
% found based $\hat{\theta}_0$. The resulting determinant is
% \begin{align}
%  \left[\det\IF^\star\right]^{-1}
%    &= \min_{\phi_u(\omega)}
%             \left[\det\IF\right]^{-1}
% \end{align}

% The optimal input can be realized as sum of $n$ sinusoids.
% Such an input is typically not robust to the initial guess since \TODO{\ldots}.
% As a solution to this, the optimal input is modified so that a fraction of the
% power in the optimal design is taken away and redistributed over all frequencies.
% This ensures that some excitation power is available also at other frequencies.

% If $\lambda_i$ is the fraction of the input power at frequency $\wn[i]$,
% then the determinant of $\IF^\star$ can be expressed as \Cite{Goodwin}
% \begin{align}
%  \det\IF^\star &= \prod_{i = 1}^{n}
%                      \lambda_i^2
%                      \alpha(\wn[i]).
% \end{align}
% The optimal input spectrum is
% \begin{align}
%  \phi_u^\star(\omega) &= \sum_{i=1}^{n}
%                            \lambda_i^2
%                            \delta(\omega,\wn[i])
% \end{align}
% If a fraction, say $\eta$, of the power is taken from each frequency and used
% at other frequencies, the input spectrum can be written as
% \begin{align}
%  \phi_u(\omega) &= (1-\eta)
%                      \sum_{i=1}^{n}
%                        \lambda_i^2 
%                        \delta(\omega,\wn[i])
%                    + \phi_u^{\robust}(\omega),\\
%                 &= (1-\eta) \phi_u^\star    (\omega)
%                    +        \phi_u^{\robust}(\omega).
% \end{align}
% The term $ \phi_u^{\robust}$ is the part that introduces robustness and
% must be such that
% \begin{align}
%  \int_0^{\pi} \phi_u^{\robust}(\omega) \dd{\omega} &= n \eta.
% \end{align}
% The spectrum $\phi_u^{\robust}(\omega)$ can for example be chosen as
% a discrete spectrum, e.g. from a multi-sine signal, or as
% a continuous spectrum. The resulting information matrix for the robust input is hence
% \begin{align}
%  \IF &= (1-\eta)\IF^\star + \IF^\robust
% \end{align}
% The determinant is therefore
% \begin{multline}
% \det\left[(1-\eta)\IF^\star + \IF^\robust\right]
%  \geq \det\left[(1-\eta)\IF^\star\right] + \det\left[\IF^\robust\right] \\
%  =  (1-\eta)^n\det\left[\IF^\star\right] + \det\left[\IF^\robust\right]
%  \text{.}
% \end{multline}
% The inequality holds since the information matrix is a positive, semi-definite
% matrix. Hence the largest possible loss in the determinant due to the robustness
% is given by the factor $(1-\eta)^n$.
% \TODO{This bound is often very low, maybe better one should be found!}
% \TODO{end}

\section{Logarithmic power distributions}\label{sec:excitation:logarithmic}
When no information about the system is known, it is desirable that the excitation signal is such that regardless of the true system characteristics, the resulting estimates have equally good statistical properties.
This can be ensured by a suitable choice of the input power spectrum.

@@

Direct calculations for $\damping < 1$ give
\begin{align}
\label{eq:excitation:auto}
\int_0^\infty\mathrm{Re}\left\{\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_1^{\HT}}{\partial \wn[1]}\right\}\frac{1}{\omega}\dd\omega = \frac{\pi}{16\wn[1]^2}\damping^{-3} + \mathcal{O}(\damping^{-1}).
\end{align}
For the corresponding element in $\partial G_{1,2}$ we start by bounding the absolute value of the integral by
% \begin{align}
% &\left|\int_0^\infty\mathrm{Re}\left\{\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_2^{\HT}}{\partial \wn[2]}\right\}\frac{1}{\omega}\dd\omega\right| \leq\label{eq:excitation:cross1} \\
% &\int_0^\infty\left|\mathrm{Re}\left\{\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_2^{\HT}}{\partial \wn[2]}\right\}\right|\frac{1}{\omega}\dd\omega \leq \\
% &\int_0^\infty\left|\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_2^{\HT}}{\partial \wn[2]}\right|\frac{1}{\omega}\dd\omega.\label{eq:excitation:cross}
% \end{align}
\begin{gather}
\left|\int_0^\infty\mathrm{Re}\left\{\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_2^{\HT}}{\partial \wn[2]}\right\}\frac{1}{\omega}\dd\omega\right|
\qquad \qquad \qquad \qquad \qquad
\label{eq:excitation:cross1} \\
\leq
\int_0^\infty\left|\mathrm{Re}\left\{\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_2^{\HT}}{\partial \wn[2]}\right\}\right|\frac{1}{\omega}\dd\omega
\leq \\
\qquad \qquad \qquad \qquad \qquad
\int_0^\infty\left|\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_2^{\HT}}{\partial \wn[2]}\right|\frac{1}{\omega}\dd\omega \text{.}\label{eq:excitation:cross}
\end{gather}
To proceed, consider
\begin{align}
\frac{\partial G_1}{\partial \wn[1]} 
  &= \frac{2\damping j\omega\left[1+\frac{j\omega}{\wn[1]\damping}\right]}
  {\wn[1]^2\left[1+2\frac{\damping j\omega}{\wn[1]} + \left(\frac{j\omega}{\wn[1]}\right)^2\right]^2},
\end{align}
which has zeros in $j\omega=\left\{0, \damping\wn[1]\right\}$. 
Hence $\left|\frac{\partial G_1}{\partial \wn[1]} \right|$ has asymptotic slope 20\unit{dB/\decade} for low frequencies, at $j\omega=\damping\wn[1]$ the slope increases to 40\unit{dB/\decade}. 
There are four poles which give a $-40$\unit{dB/\decade} roll-off after the peak. 
The shape of $\frac{\partial G_1}{\partial \wn[1]} $ is illustrated in \figref{fig:excitation:derivatives}.

@@

\section{Methodology} \label{sec:method}

  \subsection{The system under consideration}
  We consider the transfer function models of generic \gls{LTI} systems of arbitrary but finite orders $\order{B} / \order{A}$:
  \begin{align}
    G\left( \Omega, \theta \right) 
    = \frac{B\left( \Omega,\theta \right)}
                  {A\left( \Omega,\theta \right)}
    = \frac{\sum\limits_{i=0}^{\order{B}} b_i \Omega^i}
                  {\sum\limits_{i=0}^{\order{A}} a_i \Omega^i}
    \text{.} \label{eq:model}
  \end{align}

@@

  To study the behavior of one sub-system, we inspect a generic second order system $G\left( s \right)$.
  In continuous time such a system can be described by a transfer function of the form
  \begin{equation}
    G\left(s \right) = \frac{\dc}
                            {             \frac{s^2}
                                               {\wn^2} 
                             + 2 \damping \frac{s}
                                               {\wn}
                             + 1
                            }
  \text{.}
  \label{eq:secondOrderSystemCT}
  \end{equation}

  The DC gain $\dc$, the (relative) damping $\damping$ and the natural pulsation $\wn$ are a more intuitive parametrization for this system than the generic polynomial coefficients $a_i$ and $b_i$ used in \eqref{eq:model}.
  The corresponding resonance pulsation is $\wR \isdef \wn \sqrt{1-\damping^2}$.
  The $3\unit{dB}$ bandwidth around the resonance is given by $\wdB = 2\damping\wn$ for systems that have a sufficiently low damping.

  When $\damping < \frac{1}{\sqrt{2}}$, such a second order system exhibits a peak in
  the amplitude of the transfer function \citep{Oppenheim1983}:
  \begin{align}
   \left| G\left(\wM \right) \right| & \isdef
          \max_{\omega} \left| G\left( \omega \right)  \right|
        = \frac{\dc}
               {2 \damping \sqrt{1-\damping^2}}
               \label{eq:peakAmplitude}\\
   \wM &
        = \sqrt{1-2\damping^2} \wn
        \label{eq:peakFrequency}
    \text{.}
  \end{align}

@@

\subsection{Multisine Excitations}
  To excite the system under test we consider a generalization of multisine excitations \citep{Pintelon2012}:
  \begin{definition}[Generalized multisine] \label{def:generalized-MS}
  A generalized multisine $u(t)$ is a signal consisting of $F$ sine waves with a different frequencies $f_k$, amplitudes $A_k$ and phase shifts $\phi_k$:
  \begin{equation}
    u \left( t\right) = \frac{1}{\sqrt{F}}
   \sum_{k=1}^{F} 
     A_k 
     \sin 
       \left(2\pi f_k t + \phi_k \right)
  \text{.}
  \label{eq:MultisineCT}
  \end{equation}
  Without loss of generality, one can stipulate that the frequency grid is sorted, i.e. $f_1 < f_2 < \ldots < f_F$.
  \end{definition}

  For \gls{LTI} systems, it is well-known \TODO{refer to eqn. from Larsson} that the information matrix only depends upon the amplitude spectrum of the input and not upon its phase spectrum.
  As such, in this chapter assume random-phase multisines since this is a reasonable choice for systems that are dominantly linear~\TODO{referentie}.
  Hence, $\phi_k$ is the outcome of a uniform random process over $\left[0,2\pi\right[$.
  This leaves two sets of free parameters to create a signal that suits our needs:
    the amplitude of the spectral lines $A_k$ and
    the frequency grid $\left\{f_1, \ldots, f_F \right\}$.
  Note that the $k^{\text{th}}$ term in \eqref{eq:MultisineCT} is periodic with period $T_k \isdef f_k^{-1}$.

  Consequently, the complete multisine has a period $T_{u(t)}$ that is the common multiple of the period of all its components:
  \begin{equation}
    T_{u(t)}  = \lcm \set{T_1, \ldots, T_F}
      = \lcm \set{f_1^{-1}, \ldots, f_F^{-1}}
    \text{.}
  \end{equation}
  \begin{definition}[Least common (integer) multiple] \label{def:lcm-real}
  The least common multiple of the set of values $T_k \in \mathbb{F}\without\set{0}$ for $k \in \mathbb{K} \isdef \set{1, \ldots, F}$ is  
  \begin{equation}
    \lcm \set{T_1, \ldots, T_F} \isdef
      \min \Set{T |\forall k \in \mathbb{K}, \exists n_k \in \NaturalNumbersWithoutZero: n_k T_k = T }
   \end{equation}
   if such a minimum exists.
   When $\mathbb{F} \in \Set{\NaturalNumbers, \Integers, \RationalNumbers}$, a minimum always exists.
   For $\mathbb{F} = \RealNumbers$, a minimum is not guaranteed to exist (see below).
   % more info: http://mathworld.wolfram.com/LeastCommonMultiple.html
   \end{definition}
  \begin{example}
   Consider a generalized multisine where $f_1 = 1 \unit{Hz}$ and $f_2 = \tfrac{2}{3} \unit{Hz}$.
   The corresponding periods are $T_1 = f_1^{-1} = 1 \unit{s}$ and $T_2 = f_2^{-1} = 1.5 \unit{s}$.
   The overall period of the multisine is $T = n_1 T_1 = n_2 T_2 = 3 \unit{s}$, i.e. $n_1 = 3$ and $n_2=2$.
  \end{example}

@@

  This last example illustrates that a least common multiple of real numbers is not guaranteed to exist.
  Consequently, the generalized multisine of \defref{def:generalized-MS} is not guaranteed to be periodic.
  In actual implementations with limited numerical precision, this gives rise to signals that are periodic but whose period can be too long to be practically viable for periodic measurements.
  Consequently, it is undesirable to use generalized multisines that have an overly long period.

@@

  An equidistant frequency grid $\set{f_1, \ldots, f_F}$ with spacing $\Delta f$
  consisting of $F$ lines conforms to the following relation:
  \begin{equation}
    f_k = k \Delta f + k_0 \Delta f
    \emspace \forall k \in \set{1,\ldots,F},
    \emspace k_0 \in \NN
    \text{.}
  \label{eq:linGrid}
  \end{equation}

  Using such linear grid, we can easily construct a representation of a multisine
  in discrete time that is suited for use with the DFT by sampling at a given sampling rate $f_s$:
  \begin{equation}
     u \left[ n\right] = \frac{1}{\sqrt{F}}
     \sum_{k=1}^{F} 
       A_k 
       \sin 
         \left(\frac{2\pi n \left( k + k_0 \right) \Delta f}{f_s} + \phi_k \right)
    \label{eq:MultiSineDT}
  \end{equation}

@@

\subsection{(Ideal) Logarithmic Grid Multisine}
  For a logarithmic frequency grid $\set{f_1, \ldots, f_F}$ of $F$ lines,
  the following relation holds between excited frequencies:
  \begin{equation}
    f_k = \alpha \cdot f_{k-1}
    \emspace \forall k \in \set{1,\ldots,F}
  \label{eq:logGrid}
  \end{equation}
  for a given lowest frequency $f_0$ and frequency ratio $\alpha > 1$.

  When designing an excitation signal, it is common only to specify a (large) frequency band
  by its boundary frequencies $f_{\min}$ and $f_{\max}$.
  For a given $\alpha$ it is easy to determine the number of excited lines
  within the frequency range of interest:
    \begin{equation}
      F = \floor{
                  \frac{\log{f_{\max}} - \log{f_{\min}}}
                       {\log{\alpha}}
                }
    \label{eq:logms:F}
    \end{equation}
  where $\floor{\bullet}$ denotes rounding towards the nearest smaller integer value.
  In the Section \ref{sec:optimAlpha}, the choice of a suitable $\alpha$ is elaborated.

@@

  To return to the \gls{FRF}, in \secref{sec:optimAlpha} we will show that for a dense logarithmic frequency grid, any second order system with damping $\damping$ and a resonance within the bulk of the frequency grid, will receive the same number of frequency lines in its $3 \unit{dB}$ bandwidth regardless of the actual value of the resonance frequency.
  As most of the information of such systems is obtained from the measurements at frequencies inside the $3 \unit{dB}$
  bandwidth, one expects an equal variance for each system, as will be shown in \secref{sec:bestFrequencyResolution}.

@@

\subsection{Quasi-Logarithmic Grid Multisine}
  Since most measurement equipment and analysis methods rely on signals with an equidistant frequency grid, we create a frequency grid
  $\set{f_1, f_2, f_3,\ldots, f_F}$ for which \eqref{eq:logGrid} is approximately valid but with the strict constraint that the frequencies must be a subset  of an equidistant frequency grid. 
  This yields the following relation for  subsequent frequency lines:
  \begin{equation}
    f_k \approx \alpha \cdot f_{k-1}
    \emspace \forall k \in \set{1,\ldots,F}
  \label{eq:qlogGrid}
  \end{equation}
  under the constraint that
  \begin{equation}
    f_k = N \cdot \Delta f,\; N \in \NN
    \text{.}
    \label{eq:constraintLinGrid}
  \end{equation}
  Such a grid is called a quasi-logarithmic (quasi-log) grid~\citep{Pintelon2012}.

@@

  As this signal approximates a logarithmically spaced signal,  we expect that it will exhibit comparable properties.
  On the other hand, by imposing \eqref{eq:constraintLinGrid}, it remains
  possible to use signal techniques which rely on a commensurate frequency grid.

@@

  To reduce the complexity of the expressions for $\fMin{k}$ and $\fMax{k}$, we extend the frequency grid by one component to the left and to the right by means of the grid relation (equations \eqref{eq:qlogGrid} and \eqref{eq:constraintLinGrid}) to obtain the grid $\set{f_0, f_1, \ldots, f_F, f_{F+1}}$.

>>>>

\chapter{Design of excitation signals}
\label{sec:excitation}
\def\thisDir{ch02-qlogms}
\myEpigraph{A common mistake that people make when trying to design something completely foolproof is to underestimate the ingenuity of complete fools.}{Douglas Adams}{Mostly Harmless}
\researchBasedOn{Larsson2012SYSID, Geerardyn2012IMTC, Geerardyn2013TIM, }

@@

Such problem settings occur in many diverse fields of experimentation, such as the measurement of biochemical impedances~\citep{Bragos2001,Sanchez2011}, electrochemical impedance spectroscopy~\citep{Niedostatkiewicz2009,VanGheem2004}, electronics~\citep{Munir2011}, mechatronics\TODO{citation} and vibration testing\TODO{citation (ISMA?)} amongst others.

@@

  \paragraph{Contents}
  In \secref{sec:excitation:inputDesign}, we first consider (optimal) input design for an example second-order system.
  Next in \secref{}, \gls{LTI} systems are parametrized to facilitate the input design.
  This will lead to signals with a $1/f$ power spectrum as in \secref{sec:excitation:logarithmic}.
  In \secref{sec:excitation:grids} a comparison of different discrete frequency grids is provided, this is afterwards related to the problem of generating a multisine signal with a $1/f$ power spectral density.
  Next, \secref{sec:excitation:optimAlpha} describes the selection of a suitable frequency ratio $\alpha$ for a logarithmically spaced grid such that constant model quality is obtained.
  The usefulness of the proposed signals is illustrated on simulations in \secref{sec:excitation:simulation} and validated on the measurement of a set of band-pass filters in \secref{sec:excitation:measurement}.
  Conclusions are formulated in \secref{sec:excitation:conclusion}.

  \TODO{somewhere: mention optimality criterion is of lesser importance for a ``good'' signal (optimal in X = good in Y in most cases)}

\section{Input design}
\label{sec:excitation:inputDesign}
For the theoretical analysis, we consider \gls{SISO} \gls{OE} systems, i.e.
\begin{align}
y(t) &=G(s,\theta)u(t)  + e(t),
\end{align}
where $u(t)$ is a known input sequence, $y(t)$ the output sequence and $e(t)$ is zero mean Gaussian white noise.
$G$ is a rational, time invariant transfer function in continuous time. 
For notational simplicity, $s$ is used both as the complex Laplace variable and the differentiation operator; it should be clear from the context which is used. 
The transfer function is parameterized by the vector $\theta$.
These parameters are unknown and hence we are interested in estimating their values.

@@

In open loop experiments, the scaled average information matrix is given by
\begin{align}
\IF & = \lim_{N\rightarrow\infty}
          \frac{1}{N}
            \sum_{t=1}^{N}
              {\left(\frac{\partial\hat{y}(t)}{\partial\theta}\right)
               \left(\frac{\partial\hat{y}(t)}{\partial\theta}\right)^{\T}},\\
\frac{\partial\hat{y}(t)}{\partial\theta}
    & =\frac{\partial G(s,\theta)}{\partial \theta}u(t).
\end{align}
By Parseval's theorem, $\IF$ can be expressed as
\begin{align}
\IF &= \frac{1}{\pi}
         \int\limits_0^\infty \mathrm{Re}
           \left\{      \frac{\partial G(j\omega)}{\partial \theta}
                  \left[\frac{\partial G(j\omega)}{\partial \theta}\right]^{\HT}
           \right\}
           \powerspec{u}(j\omega)
           \dd{\omega} \text{.}
\label{eq:excitation:IF}
\end{align}
The information matrix is a function of the input power spectrum $\powerspec{u}$, which is the only entity that can be used to improve the quality of the estimates. 
A common optimality criterion is D-optimality where $\det{\IF}$ is maximized~\citep{Goodwin1977}.

\begin{remark}
D-optimal input designs are quite popular since this criterion minimizes the volume of  the probability density region of the parameters~\citep[Chapter 6]{Goodwin1977}.
Moreover, the D-optimal input design is independent of the parametrization of the system (as long as the different parametrizations can be related by a bijective function).
\end{remark}

@@

\section{Logarithmic power distributions}
\label{sec:excitation:logarithmic}
When no information about the system is known, it is desirable that the excitation signal is such that regardless of the true system characteristics, the resulting estimates have equally good statistical properties.
This can be ensured by a suitable choice of the input power spectrum.

@@

Direct calculations for $\damping < 1$ give
\begin{align}
\label{eq:excitation:auto}
\int_0^\infty\mathrm{Re}\left\{\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_1^{\HT}}{\partial \wn[1]}\right\}\frac{1}{\omega}\dd\omega = \frac{\pi}{16\wn[1]^2}\damping^{-3} + \mathcal{O}(\damping^{-1}).
\end{align}
For the corresponding element in $\partial G_{1,2}$ we start by bounding the absolute value of the integral by
\begin{gather}
\left|\int_0^\infty\mathrm{Re}\left\{\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_2^{\HT}}{\partial \wn[2]}\right\}\frac{1}{\omega}\dd\omega\right|
\qquad \qquad \qquad \qquad \qquad
\label{eq:excitation:cross1} \\
\leq
\int_0^\infty\left|\mathrm{Re}\left\{\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_2^{\HT}}{\partial \wn[2]}\right\}\right|\frac{1}{\omega}\dd\omega
\leq \\
\qquad \qquad \qquad \qquad \qquad
\int_0^\infty\left|\frac{\partial G_1}{\partial \wn[1]}\frac{\partial G_2^{\HT}}{\partial \wn[2]}\right|\frac{1}{\omega}\dd\omega \text{.}\label{eq:excitation:cross}
\end{gather}
To proceed, consider
\begin{align}
\frac{\partial G_1}{\partial \wn[1]} 
  &= \frac{2\damping j\omega\left[1+\frac{j\omega}{\wn[1]\damping}\right]}
  {\wn[1]^2\left[1+2\frac{\damping j\omega}{\wn[1]} + \left(\frac{j\omega}{\wn[1]}\right)^2\right]^2},
\end{align}
which has zeros in $j\omega=\left\{0, \damping\wn[1]\right\}$. 
Hence $\left|\frac{\partial G_1}{\partial \wn[1]} \right|$ has asymptotic slope 20\unit{dB/\decade} for low frequencies, at $j\omega=\damping\wn[1]$ the slope increases to 40\unit{dB/\decade}. 
There are four poles which give a $-40$\unit{dB/\decade} roll-off after the peak. 
The shape of $\frac{\partial G_1}{\partial \wn[1]} $ is illustrated in \figref{fig:excitation:derivatives}.

@@

\section{Methodology} 
\label{sec:excitation:method}

  \subsection{The system under consideration}
  We consider the transfer function models of generic \gls{LTI} systems of arbitrary but finite orders $\order{B} / \order{A}$:
  \begin{align}
    G\left( \Omega, \theta \right) 
    = \frac{B\left( \Omega,\theta \right)}
                  {A\left( \Omega,\theta \right)}
    = \frac{\sum\limits_{i=0}^{\order{B}} b_i \Omega^i}
                  {\sum\limits_{i=0}^{\order{A}} a_i \Omega^i}
    \text{.} 
    \label{eq:excitation:model}
  \end{align}

@@

  To study the behavior of one sub-system, we inspect a generic second order system $G\left( s \right)$.
  In continuous time such a system can be described by a transfer function of the form
  \begin{equation}
    G\left(s \right) = \frac{\dc}
                            {             \frac{s^2}
                                               {\wn^2} 
                             + 2 \damping \frac{s}
                                               {\wn}
                             + 1
                            }
  \text{.}
  \label{eq:excitation:secondOrderSystemCT}
  \end{equation}

  The DC gain $\dc$, the (relative) damping $\damping$ and the natural pulsation $\wn$ are a more intuitive parametrization for this system than the generic polynomial coefficients $a_i$ and $b_i$ used in \eqref{eq:excitation:model}.
  The corresponding resonance pulsation is $\wR \isdef \wn \sqrt{1-\damping^2}$.
  The $3\unit{dB}$ bandwidth around the resonance is given by $\wdB = 2\damping\wn$ for systems that have a sufficiently low damping.

  When $\damping < \frac{1}{\sqrt{2}}$, such a second order system exhibits a peak in
  the amplitude of the transfer function \citep{Oppenheim1983}:
  \begin{align}
   \left| G\left(\wM \right) \right| & \isdef
          \max_{\omega} \left| G\left( \omega \right)  \right|
        = \frac{\dc}
               {2 \damping \sqrt{1-\damping^2}}
               \label{eq:excitation:peakAmplitude}\\
   \wM &
        = \sqrt{1-2\damping^2} \wn
        \label{eq:excitation:peakFrequency}
    \text{.}
  \end{align}

@@

\subsection{Multisine Excitations}
  To excite the system under test we consider a generalization of multisine excitations \citep{Pintelon2012}:
  \begin{definition}[Generalized multisine] \label{def:excitation:generalized-MS}
  A generalized multisine $u(t)$ is a signal consisting of $F$ sine waves with a different frequencies $f_k$, amplitudes $A_k$ and phase shifts $\phi_k$:
  \begin{equation}
    u \left( t\right) = \frac{1}{\sqrt{F}}
   \sum_{k=1}^{F} 
     A_k 
     \sin 
       \left(2\pi f_k t + \phi_k \right)
  \text{.}
  \label{eq:excitation:MultisineCT}
  \end{equation}
  Without loss of generality, one can stipulate that the frequency grid is sorted, i.e. $f_1 < f_2 < \ldots < f_F$.
  \end{definition}

  For \gls{LTI} systems, it is well-known \TODO{refer to eqn. from Larsson} that the information matrix only depends upon the amplitude spectrum of the input and not upon its phase spectrum.
  As such, in this chapter assume random-phase multisines since this is a reasonable choice for systems that are dominantly linear~\TODO{referentie}.
  Hence, $\phi_k$ is the outcome of a uniform random process over $\left[0,2\pi\right[$.
  This leaves two sets of free parameters to create a signal that suits our needs:
    the amplitude of the spectral lines $A_k$ and
    the frequency grid $\left\{f_1, \ldots, f_F \right\}$.
  Note that the $k^{\text{th}}$ term in \eqref{eq:excitation:MultisineCT} is periodic with period $T_k \isdef f_k^{-1}$.

  Consequently, the complete multisine has a period $T_{u(t)}$ that is the common multiple of the period of all its components:
  \begin{equation}
    T_{u(t)}  = \lcm \set{T_1, \ldots, T_F}
      = \lcm \set{f_1^{-1}, \ldots, f_F^{-1}}
    \text{.}
  \end{equation}
  \begin{definition}[Least common (integer) multiple] \label{def:excitation:lcm-real}
  The least common multiple of the set of values $T_k \in \mathbb{F}\without\set{0}$ for $k \in \mathbb{K} \isdef \set{1, \ldots, F}$ is  
  \begin{equation}
    \lcm \set{T_1, \ldots, T_F} \isdef
      \min \Set{T |\forall k \in \mathbb{K}, \exists n_k \in \NaturalNumbersWithoutZero: n_k T_k = T }
   \end{equation}
   if such a minimum exists.
   When $\mathbb{F} \in \Set{\NaturalNumbers, \Integers, \RationalNumbers}$, a minimum always exists.
   For $\mathbb{F} = \RealNumbers$, a minimum is not guaranteed to exist (see below).
   % more info: http://mathworld.wolfram.com/LeastCommonMultiple.html
   \end{definition}
  \begin{example}
   Consider a generalized multisine where $f_1 = 1 \unit{Hz}$ and $f_2 = \tfrac{2}{3} \unit{Hz}$.
   The corresponding periods are $T_1 = f_1^{-1} = 1 \unit{s}$ and $T_2 = f_2^{-1} = 1.5 \unit{s}$.
   The overall period of the multisine is $T = n_1 T_1 = n_2 T_2 = 3 \unit{s}$, i.e. $n_1 = 3$ and $n_2=2$.
  \end{example}

@@

  This last example illustrates that a least common multiple of real numbers is not guaranteed to exist.
  Consequently, the generalized multisine of \defref{def:excitation:generalized-MS} is not guaranteed to be periodic.
  In actual implementations with limited numerical precision, this gives rise to signals that are periodic but whose period can be too long to be practically viable for periodic measurements.
  Consequently, it is undesirable to use generalized multisines that have an overly long period.

@@

  An equidistant frequency grid $\set{f_1, \ldots, f_F}$ with spacing $\Delta f$
  consisting of $F$ lines conforms to the following relation:
  \begin{equation}
    f_k = k \Delta f + k_0 \Delta f
    \emspace \forall k \in \set{1,\ldots,F},
    \emspace k_0 \in \NN
    \text{.}
  \label{eq:excitation:linGrid}
  \end{equation}

  Using such linear grid, we can easily construct a representation of a multisine
  in discrete time that is suited for use with the DFT by sampling at a given sampling rate $f_s$:
  \begin{equation}
     u \left[ n\right] = \frac{1}{\sqrt{F}}
     \sum_{k=1}^{F} 
       A_k 
       \sin 
         \left(\frac{2\pi n \left( k + k_0 \right) \Delta f}{f_s} + \phi_k \right)
    \label{eq:excitation:MultiSineDT}
  \end{equation}

@@

\subsection{(Ideal) Logarithmic Grid Multisine}
  For a logarithmic frequency grid $\set{f_1, \ldots, f_F}$ of $F$ lines,
  the following relation holds between excited frequencies:
  \begin{equation}
    f_k = \alpha \cdot f_{k-1}
    \emspace \forall k \in \set{1,\ldots,F}
  \label{eq:excitation:logGrid}
  \end{equation}
  for a given lowest frequency $f_0$ and frequency ratio $\alpha > 1$.

  When designing an excitation signal, it is common only to specify a (large) frequency band
  by its boundary frequencies $f_{\min}$ and $f_{\max}$.
  For a given $\alpha$ it is easy to determine the number of excited lines
  within the frequency range of interest:
    \begin{equation}
      F = \floor{
                  \frac{\log{f_{\max}} - \log{f_{\min}}}
                       {\log{\alpha}}
                }
    \label{eq:excitation:logms:F}
    \end{equation}
  where $\floor{\bullet}$ denotes rounding towards the nearest smaller integer value.
  In the Section \ref{sec:excitation:optimAlpha}, the choice of a suitable $\alpha$ is elaborated.

@@

  To return to the \gls{FRF}, in \secref{sec:excitation:optimAlpha} we will show that for a dense logarithmic frequency grid, any second order system with damping $\damping$ and a resonance within the bulk of the frequency grid, will receive the same number of frequency lines in its $3 \unit{dB}$ bandwidth regardless of the actual value of the resonance frequency.
  As most of the information of such systems is obtained from the measurements at frequencies inside the $3 \unit{dB}$
  bandwidth, one expects an equal variance for each system, as will be shown in \secref{sec:excitation:bestFrequencyResolution}.

@@

\subsection{Quasi-Logarithmic Grid Multisine}
  Since most measurement equipment and analysis methods rely on signals with an equidistant frequency grid, we create a frequency grid
  $\set{f_1, f_2, f_3,\ldots, f_F}$ for which \eqref{eq:excitation:logGrid} is approximately valid but with the strict constraint that the frequencies must be a subset  of an equidistant frequency grid. 
  This yields the following relation for  subsequent frequency lines:
  \begin{equation}
    f_k \approx \alpha \cdot f_{k-1}
    \emspace \forall k \in \set{1,\ldots,F}
  \label{eq:excitation:qlogGrid}
  \end{equation}
  under the constraint that
  \begin{equation}
    f_k = N \cdot \Delta f,\; N \in \NN
    \text{.}
    \label{eq:excitation:constraintLinGrid}
  \end{equation}
  Such a grid is called a quasi-logarithmic (quasi-log) grid~\citep{Pintelon2012}.

@@

  As this signal approximates a logarithmically spaced signal,  we expect that it will exhibit comparable properties.
  On the other hand, by imposing \eqref{eq:excitation:constraintLinGrid}, it remains
  possible to use signal techniques which rely on a commensurate frequency grid.

@@

  To reduce the complexity of the expressions for $\fMin{k}$ and $\fMax{k}$, we extend the frequency grid by one component to the left and to the right by means of the grid relation (equations \eqref{eq:excitation:qlogGrid} and \eqref{eq:excitation:constraintLinGrid}) to obtain the grid $\set{f_0, f_1, \ldots, f_F, f_{F+1}}$.
